Validation of Visual Statistical Inference, with Application to Linear Models

### 2012-02-02 paper_jasa.tex
up to theorem 3.1
some change

\usepackage{color}
\newcommand{\blue}[1]{{\color{blue} #1}}
\newcommand{\red}[1]{{\color{red} #1}}

@@

@@

\section{Introduction} Any statistical analysis must include some statistical graphics. For exploratory data analysis \cite{tukey:eda}, statistical graphics play an invaluable role in model checking and diagnostics. Even though we have established mathematical procedures to obtain various statistics, we need to support the results by also producing the relevant plots.

The scientific foundation of graphical methods for data analysis is well established by \cite{cleveland:1984}. In recent years we have seen several major advances in statistical graphics. Modern computing systems like R and SAS ease the production of high quality statistical plots. A grammar of graphics introduced by \cite{wilkinson:1999} presents a structured way to generate specific graphics from data and define connections between disparate types of plots.  \cite{hadley:2009} has implemented a revised version of the grammar of graphics in R, in the package {\tt ggplot2}. \blue{Graham Wills has implemented the grammar in SPSS, but I'm not sure how to cite this.}

\citet{buja:2009}, following from \cite{gelman:2004}, proposed two protocols that allow the testing and quantification of discoveries made from statistical graphics. This work represents a major advance for graphics, because it bridges the gulf between classical statistical inference procedures and exploratory data analysis.

@@

The $(m-1)$ null plots show samples drawn from the sampling distribution under $H_0$. %of the test statistic assuming that the null hypothesis is true. 
If $H_1$ is true,  the most different plot is expected to be the test statistic, i.e, the plot of the data.  A careful visual inspection follows, and the observer is asked to point out the plot most different from the lineup.  {\em If the test statistic cannot be identified} in the lineup the conclusion is to {\em not reject the null hypothesis.}

Since the lineup plot consists of $m$ plots, the probability of choosing any one of them at random is $1/m$. Thus we have a type-I error probability of $1/m$.

When a single individual identifies the observed graph in the lineup plot we report a $p$-value of at most $1/m$, otherwise the $p$-value is at least $1-\frac 1m$.

The lineup plot can be evaluated by one or more individuals.  If $N$ individuals evaluate a lineup plot independently, the number of successful evaluations, $U \sim \text{Binom} (N,\frac{1}{m})$ and we report a  $p$-value of at most $Pr(U \ge u)= \sum_{k \ge u}^N {{N \choose k} (\frac{1}{m})^k(1-\frac 1m)^{(N-k)}}$ where $u$ is the observed number of successful evaluations. %Notice that when $N=1$, this $p$-value is $\frac1m$.

\blue{'successful' evaluation - that implies that we want to reject $H_0$. We need to phrase this more carefully. }
%\subsection{Power}

\blue{This gives us a tool to evaluate different types of graphics with respect to their efficiency:} for two different visual test statistics of the same observed data, the one  is better, in which a specific pattern is more easily distinguishable visually. This should be reflected in the power of the test. We can assess power therefore both empirically based on experimental data and through theoretical considerations. Next, we will develop power theoretically and then %It is feasible to assess power theoretically and 
relate it to the empirical results.

>>>>

@@

%testing

@@

\section{Introduction} Any statistical analysis must include some statistical graphics. For exploratory data analysis, statistical graphics play an invaluable role in model checking and diagnostics. Even though we have established mathematical procedures to obtain various statistics, we need to support the results by also producing the relevant plots.

The scientific foundation of graphical methods for data analysis is well established by \cite{cleveland:1984}. In recent years we have seen several major advances in statistical graphics. Modern computing systems like R and SAS tease the production of high quality statistical plots. A grammar of graphics introduced by \cite{wilkinson:1999} presents a structured way to generate specific graphics from data and define connections between disparate types of plots.  \cite{hadley:2009} has implemented a revised version of the grammar of graphics in R, in the package {\tt ggplot2}.

\citet{buja:2009}, following from \cite{gelman:2004}, proposed two protocols that allow the testing of discoveries made from statistical graphics. This work represents a major advance for graphics, because it bridges the gulf between classical statistical inference procedures and exploratory data analysis.

@@

The $(m-1)$ null plots can be considered to be samples drawn from the sampling distribution of the the test statistic assuming that the null hypothesis is true. If $H_1$ is true, the most different plot is expected to be the test statistic, i.e, the plot of the data. {\em If the test statistic cannot be identified} in the lineup the conclusion is to {\em not reject the null hypothesis.}  A careful visual insfection follows, and the observer is asked to point out the plot most different from the lineup.

Since the lineup plot consists of $m$ plots, the probability of choosing any one of them is $1/m$. Thus we have type-I error probability of $1/m$.

The lineup plot can be evaluated by one or more individuals. When a single individual identifies the observed graph in the lineup plot we report a $p$-value of at most $1/m$, otherwise the $p$-value is at least $1-\frac 1m$.

If $N$ individuals evaluate a lineup plot independently, the number of successful evaluations, $U \sim \text{Binom} (N,\frac{1}{m})$ and we report a  $p$-value of at most $Pr(U \ge u)= \sum_{k \ge u}^N {{N \choose k} (\frac{1}{m})^k(1-\frac 1m)^{(N-k)}}$ where $u$ is the observed number of successful evaluations. %Notice that when $N=1$, this $p$-value is $\frac1m$.

%\subsection{Power}

For two different visual test statistics of the same observed data, the one  is better, in which a specific pattern is more easily distinguishable visually. This should be reflected in the power of the test. We can assess power therefore both empirically based on experimental data and through theoretical considerations. Next, we will develop power theoretically and then %It is feasible to assess power theoretically and 
relate it to the empirical results.
